Linear response theory for core set identification

### 2012-08-21 writedown/linear.response/langevin.tex
add accuracy consideration of the response theory

>>>>

\section{On the accuracy of the response fomula}

An important issue is how far can we trust the response theory.
To investigate this issue, we should firstly look at the
most important step: the trajectory measure reweighting,
i.e. Eqn.~\eqref{eqn:exact-ratio-1} and \eqref{eqn:exact-ratio-2}.
These formula are strict, however, the statistical error
may be increased do to the measure reweighting, according to
similar measure reweighting methods widely used in equilibrium sampling.
We find it more convenient to start from the nonequilibrium steady
state (NESS), then we extend it to more general, but short-time
auto-correlated dynamics.

\subsection{Starting from NESS sampling}

For the NESS system, it is possible to define a time-independent
measure of the configuration space, namely $p_0(\vect x)$, where
the subscript ``0'' denotes the reference state. We want to measure
the propties in the perturbed state, ``1'', where the NESS
probability distribtion is denoted by $p_1(\vect x)$. Assume we have
a sampling sequence $\vect x_1, \vect x_2, \cdots, \vect x_M$, which
are independent, and are subject to the reference probability distribtion
$p_0(\vect x)$. Any  observation $O$ of the perturbed system can be
estimated by
\begin{align}
  \bar O_M = \frac 1M\sum_{i=0}^{M-1} O(\vect x_i) R(\vect x_i)
\end{align}
where $R(\vect x)$ is the ration between the two probability distributions:
\begin{align}
  R(\vect x) = \frac{p_1(\vect x)}{p_0(\vect x)}
\end{align}
$\bar O_M$ is an unbiased estimator for $\langle O\rangle_1$
($\langle \cdot\rangle_1$ denotes the expectation value in the perturbed
system), because
\begin{align}
  \langle \bar O_M\rangle_0
  = \frac 1M\sum_{i=0}^{M-1}
  \int\dd \vect x\, p_0(\vect x_i)\,O(\vect x_i) R(\vect x_i)
  = \int \dd \vect x\,p_0(\vect x)\,O(\vect x)
  \,\frac{p_1(\vect x)}{p_0(\vect x)}
  = \langle O\rangle_1
\end{align}
The statistical uncertainty is defined by the variance:
\begin{align}
  \textrm{Var}_0(\bar O_M)
  = \big\langle
  [\bar O_M^2 - \langle \bar O_M\rangle_0]^2
  \big\rangle_0
  = \langle\bar O_M^2\rangle_0 -
  \langle\bar O_M\rangle_0^2 
\end{align}
It can be calculated that (only some basic algibra are needed):
\begin{align}
  \textrm{Var}_0(\bar O_M)
  =
  \langle O^2\cdot R\rangle_1 - \langle O\rangle_1^2
\end{align}
Comparing this variance with that of a direct sampling in the perturbed system,
the difference would be
\begin{align}
  \Delta\textrm{Var}_0(\bar O_M)
  =
  \langle O^2\cdot (R-1)\,\rangle_1
\end{align}
This means that the reweighted sampling can be more precise if the
observable $O$ is more significant in the region where $p_0 > p_1$, or
less precise if $O$ is more significant in the region where $p_0 <
p_1$.  This is obvious, and be observed in the equilibrium sampling
methods (for example, the umbrella sampling).  In our case, the
observable is nothing but the mollifier $K_{\sigma, \vect y}(\vect
x)$, which is the smoothified version of the delta function,
so we can somehow think
$\langle K_{\sigma, \vect y}(\vect x)\rangle_0
\approx p_0(\vect y)$ and
$\langle K_{\sigma, \vect y}(\vect x)\rangle_1
\approx p_1(\vect y)$.  This
means the average of the mollifier is always calculated with a worse
precision when $\langle K_{\sigma, \vect y}(\vect x)\rangle_1 >
\langle K_{\sigma, \vect y}(\vect x)\rangle_0$. Please note that we
are always more interested in the region where $\langle K_{\sigma,
  \vect y}(\vect x)\rangle_1 > \langle K_{\sigma, \vect y}(\vect
x)\rangle_0$, so the sad news is that we will always end up with
a worse preciseion compared with the direct sampling of perturbed
system.
To have a resonable approximation, it is required that
the unperturbed distribution $p_0(\vect x)$ should always be
overlapping with the perturbed version in the region where $p_1(\vect
x)$, or $\langle K_{\sigma, \vect y}(\vect x)\rangle_1$ is relavant.

\subsection{The accuracy of the reweighted trajectory sampling}

We want to understand what is the precision of the
reweighted trajectory sampling, namely:
\begin{align}\label{eqn:traj-ob}
  \Delta \mo (t) =
  \int \dd \vect v\, f(\vect v, 0)
  \dd \mathcal P_{i-1}^\beta[\vect X_s]\,
  O(\vect X_t)
  \Bigg[
  \frac{\dd \mathcal P^\beta_i[\vect X_s]}
  {\dd \mathcal P^\beta_{i-1}[\vect X_s]}
  \Bigg]
The leading order of the trajectory order reweighting is
\begin{align}\label{eqn:leading-rew}
  \frac{\dd \mathcal P^\beta_i[\vect X_s]}
  {\dd \mathcal P^\beta_{i-1}[\vect X_s]}
  \frac1\sigma\int_0^t M(s)\,
  \vect D(\vect X_s)\cdot \dd\vect W_t
Notice the observable $O$ only depends on the end point
of the trajectory. Therefore, we can use this properties to derive the
estimate of the path integral.
Let a series of descrete point along the trajectory:
$(\vect X_{t_1}, t_1), (\vect X_{t_2}, t_2), \cdots, (\vect X_{t_N}, t_N)$,
with $t_0 = 0,\ t_N = t$. For convenience, we denote
$\vect x_k = \vect X_{t_k}$. The descretization of the path integral
\eqref{eqn:traj-ob} is 
  \Delta\mo (t) =
  \int\dd \vect x_0\, f(\vect x_0, 0)
  \int \dd \vect x_N
  \prod_{k=0}^{N-1}
  \dd \vect x_k\,
  p(\vect x_{k+1},t_{k+1} | \vect x_{k},t_{k})\,
  O(\vect x_N)\,
  \frac1\sigma\,
  \Big[
  \sum_{m=0}^{N-1}
  M(t_m) \vect D(\vect x_m) (\vect W_{m+1} - \vect W_{m})
  \Big]
By using the Chapman-Kolmogorov property, we have
  \Delta\mo (t) = &\,
  \frac1\sigma
  \sum_{m=0}^{N-1}
  \int \dd \vect x_0\,
  \dd \vect x_m\,
  \dd \vect x_{m+1}\,
  \dd \vect x_N\,
  p(\vect x_N,t_N | \vect x_{m+1}, t_{m+1})\,
  p(\vect x_{m+1}, t_{m+1} | \vect x_{m}, t_{m})\,
  p(\vect x_{m}, t_{m} | \vect x_{0}, t_{0}) \,
  f(\vect x_0, t_0)\\
  M(t_m) \vect D(\vect x_m) (\vect W_{m+1} - \vect W_{m})  
  O(\vect x_N)
Here for simplicity, we assume the dynamics is over-damped (the equations
are the same for $O$ as under-damped dynamics).
  \dd \vect X_t =
  \vect b(\vect X_t, t)\,\dd t +
  \sigma\, \dd \vect W_t
the change of variables gives
  \vect W_{m+1} - \vect W_{m}
  \vect x_{m+1} - \vect x_{m}
  \vect b(\vect x_m, t_m)\,\Delta t_m
so
  \Delta\mo (t) = &\,
  \frac1\sigma
  \sum_{m=0}^{N-1}
  \int \dd \vect x_0\,
  \dd \vect x_m\,
  \dd \vect x_{m+1}\,
  \dd \vect x_N\,
  p(\vect x_N,t_N | \vect x_{m+1}, t_{m+1})\,
  p(\vect x_{m+1}, t_{m+1} | \vect x_{m}, t_{m})\,
  p(\vect x_{m}, t_{m} | \vect x_{0}, t_{0})\,
  f(\vect x_0, t_0)
  M(t_m) \vect D(\vect x_m) (\vect x_{m+1} - \vect x_{m})  
  O(\vect x_N)\\
  \frac1\sigma
  \sum_{m=0}^{N-1}
  \int \dd \vect x_0\,
  \dd \vect x_m\,
  \dd \vect x_N\,
  p(\vect x_N,t_N | \vect x_{m}, t_{m})\,
  p(\vect x_{m}, t_{m} | \vect x_{0}, t_{0}) \,
  f(\vect x_0, t_0)\\
  M(t_m) \vect D(\vect x_m) \vect b(\vect x_m, t_m)\,\Delta t_m
  O(\vect x_N)
% The last term equals to
%   \frac1\sigma
%   \int \dd \tau
%   M(\tau)
%   \Big\langle
%   O(\vect X_t) \vect b(\vect X_\tau,\tau) \vect D(\vect X_\tau)
%   \Big\rangle
By using the forward Kolmogorov equation, we have
  p(\vect x_{m+1},t_{m+1} | \vect x_{m},t_{m} )
  p(\vect x_{m+1},t_{m} | \vect x_{m},t_{m} )
  \Delta t_m
  \nabla_{\vect x_{m+1}}
  [ \vect b(\vect x_{m+1}, t_m)\,
  p(\vect x_{m+1},t_{m} | \vect x_{m},t_{m} )\,]\\
  \Delta t_m \frac{1} 2\,\sigma^2\,
  \nabla_{\vect x_{m+1}}^2 p(\vect x_{m+1},t_{m} | \vect x_{m},t_{m} )
with initial condition
  p(\vect x_{m+1},t_{m} | \vect x_{m},t_{m} ) =
  \delta (\vect x_{m+1} - \vect x_{m})
so
  \int\dd \vect x_{m+1}\,
  p(\vect x_N,t_N | \vect x_{m+1}, t_{m+1})\,
  p(\vect x_{m+1}, t_{m+1} | \vect x_{m}, t_{m})\,
  (\vect x_{m+1} -\vect x_{m})\\\nonumber
  \Delta t_m
  \int \dd \vect x_{m+1}\,
  p(\vect x_N,t_N | \vect x_{m+1}, t_{m+1})\,
  (\vect x_{m+1} - \vect x_{m})
  \nabla_{\vect x_{m+1}}
  [ \vect b(\vect x_{m+1}, t_m)\,
  p(\vect x_{m+1},t_{m} | \vect x_{m},t_{m} )\,]\\\nonumber
  \Delta t_m
  \int \dd \vect x_{m+1}\,
  p(\vect x_N,t_N | \vect x_{m+1}, t_{m+1})\,
  (\vect x_{m+1} - \vect x_{m})
  \frac{\sigma^2} 2\,
  \nabla_{\vect x_{m+1}}^2 p(\vect x_{m+1},t_{m} | \vect x_{m},t_{m} ) \\\nonumber
  \Delta t_m
  \int \dd \vect x_{m+1}\,
  \vect b(\vect x_{m+1}, t_m)\,
  \delta (\vect x_{m+1} - \vect x_{m})
  p(\vect x_N,t_N | \vect x_{m+1}, t_{m+1}) \\\nonumber
  \Delta t_m
  \int \dd \vect x_{m+1}\,
  \delta (\vect x_{m+1} - \vect x_{m})
  \nabla_{\vect x_{m+1}}p(\vect x_N,t_N | \vect x_{m+1}, t_{m+1})\times
  \frac{\sigma^2} 2 \\
  \Delta t_m
  \vect b(\vect x_{m}, t_m)\,
  p(\vect x_N,t_N | \vect x_{m}, t_{m+1}) 
  \,\frac{\sigma^2} 2\,\Delta t_m
  \nabla_{\vect x_{m}}p(\vect x_N,t_N | \vect x_{m}, t_{m+1})
By using the backward Kolmogorov equation
  p(\vect x_N,t_N | \vect x_{m}, t_{m+1})
  p(\vect x_N,t_N | \vect x_{m}, t_{m})
  + \mathcal O(\Delta t_m)
So by keeping the leading order of $\Delta t_m$
  \int\dd \vect x_{m+1}\,
  p(\vect x_N,t_N | \vect x_{m+1}, t_{m+1})\,
  p(\vect x_{m+1}, t_{m+1} | \vect x_{m}, t_{m})\,
  (\vect x_{m+1} -\vect x_{m})\\\nonumber
  \Delta t_m
  \vect b(\vect x_{m}, t_m)\,
  p(\vect x_N,t_N | \vect x_{m}, t_{m}) 
  \,\frac{\sigma^2} 2\,\Delta t_m
  \nabla_{\vect x_{m}}p(\vect x_N,t_N | \vect x_{m}, t_{m})
\begin{align}\label{eqn:error-o-corr}
  \Delta\mo (t) =
  \frac\sigma 2
  % \int \dd\vect x_0 f(\vect x_0,0)
  \int_0^t \dd\tau M(\tau)
  \Bigg[
  \int \dd\vect x_t\, O(\vect x_t)
  \int \dd\vect x_\tau\,
  \nabla_{\vect x_{\tau}}p(\vect x_t,t | \vect x_{\tau}, \tau)
  \vect D(\vect x_\tau)\,
  % p(\vect x_\tau,\tau | \vect x_0, 0)
  f(\vect x_\tau,\tau)
\redc{
The term in the bracket is the time correlation between
the observable and the backward flux of the error force $\vect D$.
}

\vskip .5cm
Discussions on Eqn.~\eqref{eqn:error-o-corr}:
\begin{enumerate}
\item This equation relate the error force with the error of observable,
  which would be useful in the error studies.
\item When the system does not have long-time memory:
  \begin{align}
    p(\vect x_t,t | \vect x_{\tau}, \tau) = p(\vect x_t,t ), \qquad
    \textrm{for}\ t  - \tau > t_c
  \end{align}
  Here $t_c$ denotes the length of memory of the system. Then
  Eqn.~\eqref{eqn:error-o-corr} becomes 
  \begin{align}
    \Delta\mo (t) =
    \frac\sigma 2
    % \int \dd\vect x_0 f(\vect x_0,0)
    \int_{t_c}^t \dd\tau M(\tau)
    \Bigg[
    \int \dd\vect x_t\, O(\vect x_t)
    \int \dd\vect x_\tau\,
    \nabla_{\vect x_{\tau}}p(\vect x_t,t | \vect x_{\tau}, \tau)
    \cdot
    \vect D(\vect x_\tau)\,
    % p(\vect x_\tau,\tau | \vect x_0, 0)
    f(\vect x_\tau,\tau)
    \bigg]
  \end{align}
  This means we can integrate \eqref{eqn:leading-rew} only from
  $t_c$ to $t$.
\item In our case of core set identification,
  $O(\vect x) = K_{\sigma, \vect y}(\vect x)$, which is nearly a delta function.
  A good precision of the response theory means 
  that the trajectory generated by the reference system, should
  properly sample the correlation between the
  error force flux at time $\tau$, and the probability
  density at time $t$.  In other words, we generate trajectories
  samples the conditional probability $p(\vect x_t,t | \vect x_{\tau}, \tau)$,
  and a good precision means
  the backward flux $\nabla_{\vect x_{\tau}}p(\vect x_t,t | \vect x_{\tau}, \tau)
    \cdot
    \vect D(\vect x_\tau)$ should be significant on this sampling.
\end{enumerate}
